---
title: "How to build associationsubgraphs step by step?"
# author: "Siwei Zhang"
date: "6/13/2022"
output: 
  html_document:
    toc_depth: 2
    number_sections: true
---

```{r setup, echo=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, out.width = "100%")
library(dplyr)
library(tidyverse)
library(phewasHelper)
library(associationsubgraphs)

```

<!-- # Associationsubgraphs -->

* This tutorial provides an introduction to `associationsubgraphs` package, and to conducting the complete analysis including all the steps from the structure of the input data to the final visualization using an example data set. 

# Install package and load libraries

```{r, echo=TRUE, eval=FALSE, cache=FALSE, warning=FALSE,message=FALSE}
devtools::install_github("tbilab/associationsubgraphs")

library(tidyverse)
library(associationsubgraphs)

```

# Input data

* We'll use Phecode pairs data available in the `associationsubgraphs` package as an example. The format of the input data set is similar to this data set, which is a dataframe including columns `a` and `b` representing the variables (nodes), and column `strength` that is a numeric indicator of strength of association (higher = stronger).

* Strength represents how strongly two variables are associated with each other. For instance, in this example, node pairs refer to [Phecode Pairs](https://phewascatalog.org/phecodes) where strength of the association can be measured by the odds ratio from a 2 by 2 contingency table. And please remove node pairs with `NA` missing values of `strength` in the data set.

* `associationsubgraphs` could handle large-scale input data with dimension such as the example Phecode pairs data.

```{r, echo=TRUE, cache=FALSE, warning=FALSE,message=FALSE}
#load example data set
data("phecode_pairs") 

phecode_pairs = phecode_pairs %>% 
  arrange(desc(strength)) %>% #sort the strength in descending order
  filter(!is.na(strength)) #filter out node pairs with missing values of strength

#dimension of the input data
dim(phecode_pairs) 

#overview of the phecode pairs data
head(phecode_pairs) %>% 
  knitr::kable()

```


# Annotation data

* Preparing a dataframe that has a column `id` that corresponds to the variables coded in `a` and `b` of Phecode pairs data that contains additional info of the Phecodes (nodes). For example, color and Phecode category were added to each Phecode. And the added information will be shown in the description table after clicking a subgraph to see details. 

```{r, echo=TRUE, cache=FALSE, warning=FALSE,message=FALSE}
#prepare the annotation data
annotate_node <- c(phecode_pairs$a,phecode_pairs$b) %>%
  unique() %>%
  as_tibble() %>%
  rename(id = value) %>% #rename the column corresponds to the variables to "id"
  left_join(.,phecode_def %>% dplyr::select(phecode,description,group,color) %>% dplyr::rename(id=phecode),by="id") %>% # add additional info
  arrange(group)

#overview of the annotation data
head(annotate_node) %>%
  knitr::kable()
```


<!-- * Associations of `594.10: Calculus of kidney` with other Phecodes. To explore and visualize the large-scale associations even if you have a particular interest in a node in your network, using the network that includes all combinations of pair-wise Phecodes rather than the network only includes the connection of the interested nodes with other nodes.  -->

<!-- ```{r, echo=TRUE, cache=FALSE, warning=FALSE,message=FALSE} -->
<!-- association_pairs %>% -->
<!--   filter(a == "594.10") %>% -->
<!--   head() %>%  -->
<!--   knitr::kable() -->
<!-- ``` -->

<!-- * Phecode table is shown below -->

<!-- ```{r, echo=TRUE, cache=FALSE, warning=FALSE,message=FALSE} -->
<!-- data("phecode_def")  -->

<!-- head(phecode_def) %>% -->
<!--   dplyr::select(phecode,description,group) %>%  -->
<!--   knitr::kable() -->

<!-- ``` -->

# Interactive subgraph visualization

**Calculating the subgraph structure for downstream visualization**

* We use `calculate_subgraph_structure()` to calculate subgraph structure for downstream visualization. The subgraph structure is the set of subgraphs that constructed at all strength values, and the associations were sorted in descending order of strength.

<!-- Using `calculate_subgraph_structure()` to calculate the set of subgraphs at every threshold, and the associations were sorted in descending order of strength. The nodes connected by the highest association strength are set as a cluster. The second-highest association strength is added. Specifically, if at least one node in the nodes pair with the second-highest association is shared with the first cluster that built based on the highest association, the nodes pair with the second-highest association will be added to the existing cluster including the non-shared node. Otherwise, if both nodes with the second-highest association are not shared with the first cluster, then a new separate cluster is created. This procedure is repeated for all association pairs. -->

```{r, echo=TRUE, cache=FALSE, warning=FALSE,message=FALSE}
#calculate subgraph structure
subgraphs <- phecode_pairs %>% 
  calculate_subgraph_structure()

#overview of the subgraph data
subgraphs %>% 
  dplyr::select(-subgraphs) %>%
  head() %>% 
  knitr::kable()

```


**Prepare data for downstream visualization**

* In order to present a more readable visualization, we convert Phecode to Phecode description and convert the `id` column in annotation data to Phecode description as well. When clicking the subgraph, the annotation table will show.

```{r, echo=TRUE, cache=FALSE, warning=FALSE,message=FALSE}
#convert Phecode to Phecode description
phecode_pairs = phecode_pairs %>%
  rename(phecode=a) %>%
  left_join(.,phecode_def[,c("phecode","description")],by="phecode") %>%
  rename(a=description) %>%
  dplyr::select(-phecode) %>%
  rename(phecode=b) %>%
  left_join(.,phecode_def[,c("phecode","description")],by="phecode") %>%
  rename(b=description) %>%
  dplyr::select(-phecode)

#overview of the updated phecode pairs data
phecode_pairs %>% 
  head() %>% 
  knitr::kable()

#update annotation data as well
annotate_node = annotate_node %>%
  dplyr::select(-id) %>%
  rename(id=description)
```

<!-- ## Visualizing the subgraph search -->

<!--  + The `subgraphs` includes the `rel_max_size`: size of the largest subgraph relative to the combined size of all other subgraphs, `n_subgraphs`: the number of subgraphs and `n_triples`: the number of subgraphs with at least three members. We choose the minimum `rel_max_size`, maximum `n_subgraphs` and `n_triples` to take a look. -->

<!--  + Using "largest-smallest" rule for finding the optimal threshold by tracking the minimum size of the largest subgraph relative to the combined size of all other subgraphs. -->

<!-- ```{r, echo=TRUE, cache=FALSE, warning=FALSE,message=FALSE} -->
<!-- min_rel <- subgraphs %>% -->
<!--   filter(rel_max_size == min(rel_max_size)) %>% -->
<!--   tail(1) -->

<!-- max_num_subgraphs <- subgraphs %>% -->
<!--   filter(n_subgraphs == max(n_subgraphs)) %>% -->
<!--   tail(1) -->

<!-- max_num_triples <- subgraphs %>% -->
<!--   filter(n_triples == max(n_triples)) %>% -->
<!--   tail(1) -->

<!-- subgraphs %>% -->
<!--   # filter(rel_max_size < 0.5) %>% -->
<!--   dplyr::select( -->
<!--     strength, -->
<!--     n_subgraphs, -->
<!--     max_size, -->
<!--     rel_max_size, -->
<!--     avg_density, -->
<!--     n_triples, -->
<!--     step -->
<!--   ) %>%  -->
<!--   pivot_longer(-step) %>% -->
<!--   filter(step <= 5000) %>% -->
<!--   ggplot(aes(x = step, y = value)) + -->
<!--   geom_step() + -->
<!--   geom_vline(xintercept = min_rel$step, color = 'orangered') + -->
<!--   geom_vline(xintercept = max_num_subgraphs$step, color = 'forestgreen') + -->
<!--   geom_vline(xintercept = max_num_triples$step, color = 'steelblue') + -->
<!--   facet_grid(rows = vars(name), scales = "free_y") -->


<!-- ``` -->

**Final visualization**

```{r, echo=TRUE, cache=FALSE, warning=FALSE,message=FALSE}
#visualize
visualize_subgraph_structure(
  phecode_pairs,
  node_info = annotate_node,
  subgraph_results = subgraphs,
  trim_subgraph_results = TRUE
)

```


**Highlighting a node of interest**

* If you have a particular interest in a node in your network you can "pin" that node in the visualization so the initial start point of the visualization is when that node is first added to the visible subgraphs. For instance, if you are interested in `Calculus of kidney`, simply supply the id of `"Calculus of kidney"` to the `visualize_subgraph_structure()` function and you will be automatically taken to where `Calculus of kidney` first gets grouped into a subgraph. 

```{r, echo=TRUE, cache=FALSE, warning=FALSE,message=FALSE}
#visualize
visualize_subgraph_structure(
  phecode_pairs,
  node_info = annotate_node,
  subgraph_results = subgraphs,
  trim_subgraph_results = TRUE,
  pinned_node = "Calculus of kidney"
)

```


# Generating publishable web content

* When you want to render the visualization or publish this visualization, `visualize_subgraph_structure` creates an R htmlwidget to host the visualization using `r2d3`, which means you can directly include your codes into a `.Rmd` file and then generate publishable web content by `html` file. This tool is efficient as well. For instance, total running time of calculating the subgraph structure for downstream visualization and final visualization of this Phecode pairs data on a 16 GB RAM is about 14 seconds.









